Processing genome scale tabular data with wormtable

### 2013-07-30 docs/paper/paper.tex
First full draft.

%Review style settings
%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

@@

        \paragraph*{Conclusions:} \emph{ Conclusions, brief summary and potential implications. }
\end{abstract}

@@

Another problem with tables stored text files is that it is difficult 
to index the information within particular columns. 
This means that most operations on the table require a complete scan 
through the file. As a result, the only viable means of 
working with a subset of the data
is to create another file consisting of the subset of interest. 
This is inflexible and error prone, and multiplies already significant  
storage requirements.

@@

% what is wormtable?
Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data of essentially unlimited size. Tables can
be either written directly via the library, or converted
from existing formats via command line tools. 
Wormtable provides a simple Python API to access rows
in the table, facilitating efficient and convenient access to 
data.

% Wormtable's binary row format 
Data is stored using a  
compact binary format designed specifically for biological applications.
Each row consists of a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Wormtable currently supports signed and unsigned integers, 
floating point and character data. To allow data to be stored as 
efficiently as possible, wormtable supports a wide range of 
type sizes: integers range from one to eight bytes long,
and IEEE half, single and double precision floating point 
are supported. Fixed and variable length character strings 
are also supported.

% Tables

% why do we use Berkeley DB?
To take advantage of the data processing capabilities of database
technologies without the overheads and disadvantages of relational databases,
wormtable uses Berkeley DB~\cite{obs99} for row storage and indexing.
Berkeley DB is a free and open-source embedded database toolkit that provides a
scalable key-value store. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\cite{sb12}. Berkeley DB is a highly modular library, providing the 
means to define a very lightweight read-only database, without the 
overheads and complexity of a relational database system.

% Indexes

In wormtable, each table corresponds to a directory in the file system
which is used to store the table, indexes and metadata.
Indexes over arbitrary combinations of columns are supported, 
and these are managed using the \texttt{wtadmin} program, which 
provides several other administration utilities.
Wormtable files are portable across operating systems 
and processor architectures.

% Layout & implementation.
Each table corresponds to a single directory in the file system.

@@

\subsection{API}

\subsection*{Examples}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}.  
To illustrate the advantages of wormtable, we
converted a large 
publicly available VCF file, produced as part of the
Drosophila genetic reference panel~\cite{mr+12}, to wormtable format.
This VCF\cite{freeze.vcf}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a standard workstation with a single hard disk. The
conversion required approximately 68 minutes, and the size of the resulting  
wormtable file was 10.1GB (using the smallest type required
to represent the data in each column).

Values are stored in wormtable in a portable binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset using wormtable and PyVCF~\cite{pyvcf}
a Python VCF parser.
In this example we proceed 
row-by-row, examining the REF and ALT columns and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 126 minutes;
the wormtable version required
55 seconds. To compare against methods that are known to be extremely
efficient, we repeated the same 
example using the Unix tools \texttt{cut}, \texttt{grep} 
and \texttt{awk},  which required 77 seconds. Both the
Unix pipeline and wormtable were limited by the sequential read
bandwidth of the hard drive, and would therefore be much 
faster using modern solid state storage.
It should be noted, however, that this simple task represents 
something of a best case scenario for the Unix tools pipeline.

Another major advantage of wormtable is the ability to create 
an index over arbitrary combinations of columns within a table.
Indexes make many operations much more efficient, as they provide 
a way to quickly find particular rows and avoid the 
heavy cost of a full table scan.
To illustrate this advantage, we repeated the example 
of counting transitions and transversions using an index 
on the REF and ALT columns.
The index required 3m42s to build and 
consumed 66MB of storage space. Counting the number of 
transitions and transversions using this index required 
less than a second.

\section{Discussion}
It is well known that parsing text files is a very inefficient process, 
and binary file formats are preferable for this reason. The BCF format,
for example, is a binary version of the VCF format discussed above which 
is designed to avoid just this problem. There are significant difficulties,
however, with having many different binary file formats for bioinformatics 
data. Each binary format requires a library and set of tools to view 
and process it, as it is not reasonable to expect users to decode binary
files. Bindings for several different languages must also be provided, if 
the file format is to be widely used.
Maintaining these libraries, tools and language bindings across 
different processor architectures and operating systems
is a complex software engineering task. Maintaining this ecosystem 
separately for several different file formats is surely 
unsustainable.

Wormtable alleviates the need for these different formats and libraries,
as it is flexible enough to store many different types of data. Python 
is currently the only programming language supported, but a C library
and bindings for other languages are planned. Wormtable is portable,
and has been tested on big- and little-endian platforms with 32 and 64
bit word sizes, along with many operating system combinations. To 
take advantage of the advanced data processing features of wormtable
all that is required is a conversion program, a considerably simpler
task than designing and supporting a custom binary file format.

@@

The library supports efficient access to any data stored in 
wormtable format, but there is currently a conversion tool for
the VCF format only. The most important aspect of future development 
is to develop tools to convert other tabular formats such as 
PSL, GFF, and SAM to wormtable format [Dan, Rob: does this make sense?
are there other formats we should chuck in here?]. Since the conversion 
tool can be written entirely in Python this is not a difficult task.

Other probable directions for future development of wormtable would
be to incorporate BGZF compression~\cite{li11}, allowing for smaller
files with efficient random access; to develop a 
C library and bindings for other popular languages such as 
Perl and R; and to generally enrich the API with features useful 
to the community. 
Wormtable is an open and collaborative project, which would welcome 
feedback and contributions from the community

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}

\emph{This should state clearly the main conclusions of the article and give a clear 
explanation of the importance and relevance of the software.}

>>>>

%Review style settings
\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

@@

\paragraph*{Conclusions:} Using wormtable's simple API users 
can process tables in a very high performance manner, allowing them
much greater freedom to explore their data. 
Since it is no longer necessary to parse complex files,
data analysis code is also much simpler and less error prone. 
\end{abstract}

@@

Another problem with tables stored text files is that it is difficult 
to index the information within particular columns. 
This means that most operations on a table require a complete scan 
through the file. As a result, the only viable means of 
working with a subset of the data
is to create another file consisting of the subset of interest. 
This is inflexible and error prone, and multiplies already significant  
storage requirements.

@@

% what is wormtable?
Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data of essentially unlimited size. Tables can
be either written directly via the library, or converted
from existing formats via command line tools. 
Wormtable provides a simple Python API to access rows
in the table, facilitating efficient and convenient access to 
data. The core data manipulation and storage facilities 
are written in C for efficiency.

% Wormtable's binary row format 
Data is stored in rows using a  
compact binary format designed specifically for biological applications.
Each row consists of a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Wormtable currently supports signed and unsigned integers, 
floating point and character data. To store data as 
compactly as possible, integer sizes range from one to eight bytes,
and real numbers are stored as IEEE half, single, and double 
precision floating point values. 
Fixed and variable length character strings 
are also supported.

% Tables
Rows are stored sequentially in a data file, and the offset and length 
of each row is stored in a Berkeley DB~\cite{obs99} database. 
Thus, any given row can be found 
by first looking up the database for its offset and length, and by then reading
the required set of bytes from the data file. 
Berkeley DB is a free and open-source embedded database toolkit that provides a
scalable key-value store. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\cite{sb12}.

% Indexes
Besides random access to rows, wormtable also provides indexes
over arbitrary combinations of columns. An index is a database in 
which the keys are the values from the columns in question concatenated
together, and the values are pointers to the original row. Indexes make
many operations much more efficient, since we can directly seek to the 
rows we are interested in without performing a full table scan. They also 
provide a very efficient means of calculating a histogram for a given 
set of columns.

% Layout 
Each table corresponds to a directory in the file system
which is used to store the data files, indexes and metadata.
Files within a table's home directory are not intended to be 
manipulated by users, and the \texttt{wtadmin} program is provided 
to perform administration tasks such as adding and removing 
indexes. This approach is 
very flexible, since no centralised storage is required and 
tables can be moved around and between systems at will.
Tables are portable across operating systems and hardware
architectures.

@@

\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using a simple interface. The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read (similar to 
a select list in SQL). Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings.
Providing an index as an argument to the \texttt{cursor} function
results in the rows being visited in the order defined by the 
index. The arguments \texttt{start} and \texttt{stop} 
then stipulate the range of values we are interested in.
There are several other useful classes and methods, but 
the \texttt{cursor} method encapsulates the core functionality
of wormtable: efficient iteration over rows.

The interface is kept as
simple and similar to the Python standard library as possible,
to ensure that there is little difficulty in learning the API.
Similarly, although one can readily imagine defining a 
query language using these 
primitive operations, we have deliberately avoided doing so. Adding 
an SQL-like query language would greatly increase the complexity 
of wormtable, and would only serve to raise the barrier to entry.
Less experienced programmers 
do not need to learn a query language to process data 
in a very high performance manner.

\subsection*{Examples}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}.  
To illustrate the advantages of wormtable, we
converted a large 
publicly available VCF file, produced as part of the
Drosophila genetic reference panel~\cite{mr+12}, to wormtable format.
This VCF\cite{freeze.vcf}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a standard workstation with a single hard disk. The
conversion required approximately 69 minutes, and the size of the resulting  
wormtable was 10GB (using the smallest type required
to represent the data in each column).

Values are stored in wormtable in a portable binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset using wormtable and PyVCF~\cite{pyvcf}
a Python VCF parser.
In this example we proceed 
row-by-row, examining the REF and ALT columns and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 126 minutes;
the wormtable version required
57 seconds. To compare against methods that are known to be extremely
efficient, we repeated the same 
example using the Unix tools \texttt{cut}, \texttt{grep} 
and \texttt{awk},  which required 80 seconds. Both the
Unix pipeline and wormtable were limited by the sequential read
bandwidth of the hard drive, and would therefore be much 
faster using modern solid state storage.
It should be noted, however, that although Unix 
pipelines are efficient, it is a difficult and error prone method 
of processing data when more complex calculations are required.

Another advantage of wormtable is the ability to create 
an index over arbitrary combinations of columns within a table.
Indexes make many operations much more efficient, as they provide 
a way to quickly find particular rows and avoid the 
heavy cost of a full table scan.
To illustrate this advantage, we repeated the example 
of counting transitions and transversions using an index 
on the REF and ALT columns.
The index required 3m40s to build and 
consumed 66MB of storage space. Counting the number of 
transitions and transversions using this index required 
less than a second.

\section*{Discussion}
It is well known that parsing text files is a very inefficient process, 
and binary file formats are preferable for this reason. BCF,
for example, is a binary version of the VCF format discussed above which 
is designed to avoid just this problem. There are significant difficulties,
however, with having many different binary file formats for bioinformatics 
data. Each binary format requires a library and set of tools to view 
and process it, as it is not reasonable to expect users to decode binary
files. Bindings for several different languages must also be provided, if 
the file format is to be widely used.
Maintaining these libraries, tools and language bindings across 
different processor architectures and operating systems
is a complex software engineering task. Maintaining this ecosystem 
separately for several different file formats is surely 
unsustainable.

Wormtable alleviates the need for these different formats and libraries,
as it is flexible enough to store many different types of data. Wormtable is portable,
and has been tested on big- and little-endian platforms with 32 and 64
bit word sizes, along with many operating system combinations. To 
take advantage of the advanced data processing features of wormtable
all that is required is a conversion program, a considerably simpler
task than designing and supporting a custom binary file format.

The library supports efficient access to any data stored in 
wormtable format, but there is currently a conversion tool for
the VCF format only. The most important aspect of future development 
is to develop tools to convert other tabular formats such as 
PSL, GFF, and SAM to wormtable format [Dan, Rob: does this make sense?
are there other formats we should chuck in here?].

Other probable directions for future development of wormtable 
are to incorporate data file compression; to develop a 
C library and bindings for other popular languages such as 
Perl and R; and to generally enrich the API with features useful 
to the community. It is difficult to predict the 
precise direction of future development since this depends 
on community involvement.  Wormtable is an open and 
collaborative project, actively seeking feedback and 
contributors.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
\emph{This should state clearly the main conclusions of the article and give a clear 
explanation of the importance and relevance of the software.}

The volumes of data being produced and processed in bioinformatics are constantly 
increasing, but the tools available to end users to 
process this data are still mostly based on parsing plain text. This 
approach is very inefficient, and leads to several undesirable outcomes.
Firstly, and most obviously, a researcher's productivity is inevitably
reduced by having to wait for several hours to see the result of a 
simple calculation. There are also some less obvious consequences:
code quality is reduced, since testing over the entire dataset is 
infeasible, and it is much less likely that the effects of 
changing arbitrary quality thresholds (for example) will be 
systematically examined.

@@
